---
title: "Conjugate models"
subtitle: "Exercise BADT"
author: "Sahlin"
---

### Background 

Conjugate models are very useful even today, but for relatively simple problems. 

### Task 1

Test an hypothesis using conjugate model. How sensitive is the conclusion from the choice of prior? 


Let $\lambda$ be the average number of goals scored in a Women’s World Cup game. We’ll analyse $\lambda$ by a Gamma-Poisson model where data  $Y_i$ is the observed number of goals scored in a sample of World Cup games

1. Specify the model 

2. Plot and summarize our prior understanding of  $\lambda$.

3. Why is the Poisson model a reasonable choice for our data  
$Y_i$?

Use the wwc_2019_matches data from [fivethirtyeight](https://data.fivethirtyeight.com/)

The wwc_2019_matches data includes the number of goals scored by the two teams in each 2019 Women’s World Cup match. We summed the scores by the two teams per game, made a histogram and calculated some summary statistics:

```{r}
#| echo: false

library(fivethirtyeight)

goals <- wwc_2019_matches$score1 + wwc_2019_matches$score2

library(ggplot2)

hist(goals)
```


```{r}
#| echo: false

summary(goals)
length(goals)
```

4. Make a test if the average number of goals scored in a Women’s World Cup game is less than 1.8.


### Task 2

Use conjugate models to construct probability intervals for the results from a clinical trial and compare to a published meta-analysis. 

[Lancet paper](https://www.thelancet.com/article/S0140-67362031142-9/fulltext)


![](https://www.thelancet.com/cms/10.1016/S0140-6736(20)31142-9/asset/b966c770-af7d-4424-a04a-556cef203d13/main.assets/gr4_lrg.jpg)

A relative risk (RR) is a ratio of the probability of an event occurring in the exposed group versus the probability of the event occurring in the non-exposed group.

1. Reproduce the probability interval for a RR from the forest plots tables e.g. Scales et al 2003

2. Calculate a probability interval for one of the studies for which the RR is not calculable, e.g. Hall et al 2014



 
